---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
1. The following code loads the data into the **activity** dataframe.  No data transformations are needed.
```{r}
library(dplyr, warn.conflicts=FALSE)
library(ggplot2)
activity <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
1. This histogram plots the frequency of total number of steps taken each day.  Per the instructions, missing values in **activity** are ignored.
```{r}
stepsByDate <- na.omit(activity) %>%
    group_by(date) %>%
    summarise(totsteps=sum(steps))
ggplot(data=stepsByDate, aes(x=totsteps)) +
    geom_histogram() +
    ggtitle("Frequency of Steps per Day") +
    xlab("steps / day")
```

## What is the average daily activity pattern?



## Imputing missing values
1. The following calculates the total number of rows in **activity** containing NAs ...
```{r}
sum(is.na(activity$step))
```


## Are there differences in activity patterns between weekdays and weekends?
